---
title: "Class 9: Candy Mini-Project"
author: "Brooke Clements (PID_A17532793)"
format: pdf
prefer-html: true
---

## Exploratory Analysis of Halloween Candy

```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy=read.csv(candy_file, row.names=1)
head(candy)
```

## What is in the dataset? 
> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

## What is your favorite candy? 

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Twix") |> 
  select(winpercent)
```
> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
candy |> 
  filter(row.names(candy)=="100 Grand") |> 
  select(winpercent)
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy |> 
  filter(row.names(candy)=="Kit Kat") |> 
  select(winpercent)
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy |> 
  filter(row.names(candy)== "Tootsie Roll Snack Bars") |> 
  select(winpercent)
```

`skim()` This function can give us a quick overview of the dataset. 

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent variable looks to be the most different from the rest of the dataset. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

A 0 means that there is chocolate in the candy while a 1 means that there is no chocolate in the candy.  

## Exploratory analysis

> Q8. Plot a histogram of winpercent values using both base R an ggplot2.

```{r}
library(ggplot2)

ggplot(candy, aes(x=winpercent)) +
  geom_histogram()

```

> Q9. Is the distribution of winpercent values symmetrical?

No, this distribution of winpercent values is not symmetrical.

> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])
```
Chocolate on average is ranked higher than fruity candy. 

> Q12. Is this difference statistically significant?

```{r}
t.test(candy$winpercent[candy$chocolate==1], candy$winpercent[candy$fruity==1])
```
The P-value is less than 0.05 meaning that this data is statistically significant. 

## Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent, decreasing = TRUE),], n=5)
```
>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy)+
  aes(winpercent, rownames(candy))+
  geom_bar(stat="identity")+
  theme(axis.text.y = element_text(size=3))
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy)+
  aes(winpercent, reorder(rownames(candy),winpercent))+
  geom_bar(stat="identity")+
  theme(axis.text.y = element_text(size=3))
```

## Time to add some useful color
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```
```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)+
  theme(axis.text.y = element_text(size=3))
```

> Q17. What is the worst ranked chocolate candy?

Sixlets is the worst ranked chocolate candy. 

> Q18. What is the best ranked fruity candy?

Nik L Nip is the worst ranked fruity candy.

## Taking a look at pricepercent

```{r}
library(ggrepel)
ggplot(candy)+
  aes(x=winpercent, y=pricepercent, label=rownames(candy))+
  geom_point(col=my_cols)+
  geom_text_repel(col=my_cols, size=3.3, max.overlaps=5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

## Exploring the correlation structure

```{r}
library(corrplot)
```
```{r}
cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolate have strong negative correlation. 

>Q23. Similarly, what two variables are most positively correlated?

Winpercent and chocolate have the most positive correlation.

## Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
```{r}
plot(pca$x [,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x= PC1, y= PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
ggplotly(p)
```

```{r}
ggplot(pca$rotation) +
  aes(PC1, reorder(row.names(pca$rotation),PC1)) +
  geom_col()
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

Fruity candy is shown in this grapg to be strongly positive which can be also seen in the corrplot.



