---
title: "Class14"
author: "Brooke Clements (PID: 17532793)"
format: pdf
---

```{r}
#| include: false
library(DESeq2)
```



## Section 1. Differential Expression Analysis

```{r}
metaFile <- "GSE37704_metadata - GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts - GSE37704_featurecounts.csv"
```

```{r}
countData <- read.csv("GSE37704_featurecounts - GSE37704_featurecounts.csv")
colData <- read.csv("GSE37704_metadata - GSE37704_metadata.csv")
```


```{r}
# Import metadata and take a peek
colData = read.csv(metaFile, row.names=1)
head(colData)
```

```{r}
# Import countdata
countData = read.csv(countFile, row.names=1)
head(countData)
```

> Q. Complete the code below to remove the troublesome first column from countData

```{r}
# Note we need to remove the odd first $length col
countData <- as.matrix(countData[,-1])
head(countData)
```

> Q. Complete the code below to filter countData to exclude genes (i.e. rows) where we have 0 read count across all samples (i.e. columns).

```{r}
# Filter count data where you have 0 read count across all samples.
countData = countData[rowSums(countData) > 0, ]
head(countData)
```

# Running DESeq2

```{r}
dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~ condition)
dds = DESeq(dds)
dds
```

```{r}
res = results(dds)
```

>Q. Call the summary() function on your results to get a sense of how many genes are up or down-regulated at the default 0.1 p-value cutoff.

```{r}
summary(res)
```

## Volcano Plot

```{r}
library(ggplot2)

ggplot(res) +
  aes(x= log2FoldChange, y= -log10(padj)) +
  geom_point()
```

> Q. Improve this plot by completing the below code, which adds color, axis labels and cutoff lines:

```{r}
# Make a color vector for all genes
mycols <- rep("gray", nrow(res) )

# Color blue the genes with fold change above 2
mycols[ abs(abs(res$log2FoldChange)) > 2 ] <- "blue"

# Color gray those with adjusted p-value more than 0.01
mycols[ res$padj > 0.01 ] <- "gray"

ggplot(res) +
  aes(x= log2FoldChange, y= -log10(padj)) +
  geom_point(col = mycols) +
  xlab("Log2(FoldChange)") +
  ylab("-Log(P-value)") +
  geom_vline(xintercept  = c(-2,2)) +
  geom_hline(yintercept = -log10(0.01))
```

## Adding gene annotation

> Q. Use the mapIDs() function multiple times to add SYMBOL, ENTREZID and GENENAME annotation to our results by completing the code below.

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res), 
                    keytype="ENSEMBL",
                    column="SYMBOL",
                    multiVals="first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="GENENAME",
                    multiVals="first")

head(res, 10)
```

> Q. Finally for this section let's reorder these results by adjusted p-value and save them to a CSV file in your current project directory.

```{r}
res = res[order(res$pvalue),]
write.csv(res, file="deseq_results.csv")
```

## Section 2. Pathway Analysis

```{r}
library(pathview)
library(gage)
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

```{r}
attributes(keggres)
```

```{r}
# Look at the first few down (less) pathways
head(keggres$less)
```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```

```{r}
# A different PDF based output of the same data
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
```

```{r}
## Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```

```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```

> Q. Can you do the same procedure as above to plot the pathview figures for the top 5 down-regulated pathways?

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04640")

```
```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04630")

```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa00140")

```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04142")

```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04330")

```

## Section 3. Gene Ontology (GO)

```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets)

lapply(gobpres, head)
```

## Section 4. Reactome Analysis

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

> Q: What pathway has the most significant “Entities p-value”? Do the most significant pathways listed match your previous KEGG results? What factors could cause differences between the two methods?

Cell Cycle, Mitotic has the most significant “Entities p-value”. M phase of mitotic cell cycle is the previous KEGG results and it matches. 




