---
title: "Class05"
author: "Brooke Clement (PID:A17532793)"
format: html
---

# Class 5 Data Visualization Lab

#Install the package ggplot2
#install.packages("ggplot2")


# Any time I want to use this package I need to load it

```{r}
library(ggplot2)
```

```{r}
View(cars)
```


# a quick base R plot - this is not ggplot
```{r}
plot(cars)
```


# Our first ggplot
# We need data + aes + geoms
```{r}
ggplot(data=cars) +
  aes(x=speed, y=dist) +
  geom_point()

p <- ggplot(data=cars) +
  aes(x=speed, y=dist) +
  geom_point()
```


# Add a line geom with geom_line()
```{r}
p + geom_line()
```

# Add a trendline close to the data
```{r}
p + geom_smooth(method="lm")
```

p + geom_smooth()


#____________________________#

# Read in our drug expression data
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```


# Let's make our first plot attempt
```{r}
g <- ggplot(data=genes) +
  aes(x=Condition1, y=Condition2,
      col=State) +
  geom_point()

```


# Q. How many genes are in this dataset?
```{r}
nrow(genes)
```


# Q. How many columns did you find?
```{r}
ncol(genes)
```


# Q. How many ‘up’ regulated genes there are.
```{r}
table(genes$State)
```


# Q. What fraction of total genes is up-regulated in this dataset?
```{r}
round((table(genes$State) /nrow(genes))*100, 2)
```


# Add some color
```{r}
g + scale_color_manual(values=c("blue", "gray","purple")) +
  labs(title="Gene expression changes upon drug treatment",
    x="Control(no drug)",
    y="Drug Treatment") +
  theme_bw()
```



