---
title: "Class 18: Pertussis and the CMI-PB project"
author: "Brooke Clements (PID:17532793)"
format: pdf
---

# Background
Pertussis (whooping cough) is a common lung infection caused by the bacteria B. Pertussis. It can infect anyone but is most deadly for infants (under 1 year of age). 

# CDC tracking data
The CDC tracks the number of pertussis cases: https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html


```{r, echo=FALSE}
cdc <- data.frame(
  year = c(
    1922L,
    1923L,
    1924L,
    1925L,
    1926L,
    1927L,
    1928L,
    1929L,
    1930L,
    1931L,
    1932L,
    1933L,
    1934L,
    1935L,
    1936L,
    1937L,
    1938L,
    1939L,
    1940L,
    1941L,
    1942L,
    1943L,
    1944L,
    1945L,
    1946L,
    1947L,
    1948L,
    1949L,
    1950L,
    1951L,
    1952L,
    1953L,
    1954L,
    1955L,
    1956L,
    1957L,
    1958L,
    1959L,
    1960L,
    1961L,
    1962L,
    1963L,
    1964L,
    1965L,
    1966L,
    1967L,
    1968L,
    1969L,
    1970L,
    1971L,
    1972L,
    1973L,
    1974L,
    1975L,
    1976L,
    1977L,
    1978L,
    1979L,
    1980L,
    1981L,
    1982L,
    1983L,
    1984L,
    1985L,
    1986L,
    1987L,
    1988L,
    1989L,
    1990L,
    1991L,
    1992L,
    1993L,
    1994L,
    1995L,
    1996L,
    1997L,
    1998L,
    1999L,
    2000L,
    2001L,
    2002L,
    2003L,
    2004L,
    2005L,
    2006L,
    2007L,
    2008L,
    2009L,
    2010L,
    2011L,
    2012L,
    2013L,
    2014L,
    2015L,
    2016L,
    2017L,
    2018L,
    2019L,
    2020L,
    2021L,
    2022L,
    2023L,
    2024L,
    2025L
  ),
  cases = c(
    107473,
    164191,
    165418,
    152003,
    202210,
    181411,
    161799,
    197371,
    166914,
    172559,
    215343,
    179135,
    265269,
    180518,
    147237,
    214652,
    227319,
    103188,
    183866,
    222202,
    191383,
    191890,
    109873,
    133792,
    109860,
    156517,
    74715,
    69479,
    120718,
    68687,
    45030,
    37129,
    60886,
    62786,
    31732,
    28295,
    32148,
    40005,
    14809,
    11468,
    17749,
    17135,
    13005,
    6799,
    7717,
    9718,
    4810,
    3285,
    4249,
    3036,
    3287,
    1759,
    2402,
    1738,
    1010,
    2177,
    2063,
    1623,
    1730,
    1248,
    1895,
    2463,
    2276,
    3589,
    4195,
    2823,
    3450,
    4157,
    4570,
    2719,
    4083,
    6586,
    4617,
    5137,
    7796,
    6564,
    7405,
    7298,
    7867,
    7580,
    9771,
    11647,
    25827,
    25616,
    15632,
    10454,
    13278,
    16858,
    27550,
    18719,
    48277,
    28639,
    32971,
    20762,
    17972,
    18975,
    15609,
    18617,
    6124,
    2116,
    3044,
    7063,
    22538,
    21996
  )
       )
```


I want a plot if year vs cases
```{r}
library(ggplot2)

ggplot(cdc)+
  aes(year, cases)+
  geom_point() +
  geom_line()
```

> Q. Add anotation lines for the major milestones of wP vaccination roll-out (1946) and the switch to the aP vaccine (1996). 

```{r}
ggplot(cdc)+
  aes(year, cases)+
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, col="blue", lty=2)+
  geom_vline(xintercept = 1996, col="purple", lty=2) +
  geom_vline(xintercept = 2020, col ="darkgreen", lty=2) 
```

After 1946 (wP vaccination roll-out) we see the the cases of Pertussis starts to decrease dramatically.Then in 1946 (switch to the aP vaccine) we start to see an increase in cases. This could be due to a few reasons for this. One could be an increase in people getting testing, as well as more sensitive PCR-based testing. Another reason could be that the aP vaccine has a smaller efficacy when compared to the wP vaccine. This means that after about 12 years the aP vaccine will need a booster shot while the wP vaccine does not. 

# Exploring CMI-PB data
The CMI-PM project < https://www.cmi-pb.org/ > mission of CMI-PB is to provide the scientific community with a comprehensive, high-quality and freely accessible resource of Pertussis booster vaccination.

Basically, make avalable a large dataset on the immune response to Pertussis. They use a "booster" vaccination as a proxy for Pertussis infection. 

They make their data avalable as JSON format API. We can read this into R with the `read_json()` function from the **jsonlite** package: 

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)
head(subject)
```

> Q. How many aP and wP individuals are there in this `subject` table?

```{r}
table(subject$infancy_vac)
```

> Q. How many male/female are there?

```{r}
table(subject$biological_sex)
```

> Q. What is the break down of `biological_sex` and `race`?

```{r}
table(subject$race, subject$biological_sex)
```

> Q. Is this representative of the US population?

This data is not representative of the US population. This study was done during the 2020 lockdown and this data is more represanative of the UCSD population.

We can read more tables from CMI-PB database

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE)
ab_titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE)
```

```{r}
head(specimen)
```

```{r}
head(ab_titer)
```

To make sense of all this data we need to "join" (a.k.a. "merge" or "link") all thses tables together. Only then will you know that a given Ab measurement (from the `ab_titer` table) was collected on a certain date (from the `specimen` table) from a certain wP or aP subject (from the `subject` tabel).

We can is **dplyr** and the `*join()` family of function to do this. 
```{r}
library(dplyr)
meta <- inner_join(subject, specimen) 
head(meta)
```

Let's do one more `inner_join()` to join the `ab_titer` with all our `meta` data. 

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```

> Q. How many different Ab "isotype" values are in this dataset?

```{r}
table(abdata$isotype)
```


> Q. How many different "antigen" values are in this dataset?

```{r}
table(abdata$antigen)
```

Let's focas on IgG isotype

```{r}
igg <- abdata |>
  filter(isotype=="IgG")
head(igg)
```

Let's make a plot of `MFI_normalised` values for all `antigen`values. 

```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen)+
  geom_boxplot()
```

The antigen "PT","FIM2/3", and "FHA" appear to have the widest range of values. 

> Q. Is there a difference for these responses between aP and wP individuals? 

```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen, col=infancy_vac)+
  geom_boxplot()
```

```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen)+
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```

> Q. Is there a difference with time (i.e before booster shot vs after booster shot)?

```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen, col=infancy_vac)+
  geom_boxplot() +
  facet_wrap(~visit)
```

```{r}
abdata.21 <- abdata |> filter(dataset=="2021_dataset")
abdata.21 |> filter(isotype =="IgG", antigen =="PT") |>
  ggplot()+
  aes(x=planned_day_relative_to_boost, y=MFI_normalised,col=infancy_vac, group=subject_id)+
  geom_point() +
    geom_line() +
    geom_smooth(aes(group = infancy_vac), se =FALSE, linewidth=2, span =0.3) +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

# Obtaining CMI-PB RNASeq data

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```

```{r}
#meta <- inner_join(specimen, subject)
ssrna <- inner_join(rna, meta)
```

> Q. Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of `visit` vs. `tpm`).

```{r}
ggplot(ssrna) +
  aes(visit, tpm, group=subject_id) +
  geom_point() +
  geom_line(alpha=0.2)
```

> Q.What do you notice about the expression of this gene (i.e. when is it at it’s maximum level)?

> Q. Does this pattern in time match the trend of antibody titer data? If not, why not?

```{r}
ggplot(ssrna) +
  aes(tpm, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(vars(visit))
```

```{r}
ssrna %>%  
  filter(visit==4) %>% 
  ggplot() +
    aes(tpm, col=infancy_vac) + geom_density() + 
    geom_rug() 
```




